### 2018-05-08 doc/paper.tex
Another "review" pass.

The paper is organized as follows. Section~\ref{sec:notation} introduces some
basic notation and results about floating point arithmetic, error-free
transformations and the de Casteljau algorithm. In
Section~\ref{sec:compensated-2},
the compensated algorithm for polynomial evaluation from \cite{Jiang2010} is
reviewed and notation is established for the expansion. In
Section~\ref{sec:compensated-k}, the \(K\) compensated algorithm is provided and
a forward error analysis is performed. Finally, in Section~\ref{sec:numerical} we
give numerical tests to illustrate the practical efficiency of our algorithms.

@@

An error-free transformation is a computational method where both
the computed result and the round-off error are returned. It
is considered ``free'' of error if the round-off can be represented
exactly as an element of \(\mathbf{F}\).
The error-free transformations used in this paper are
the \texttt{TwoSum} algorithm by Knuth \cite{Knuth1997} and \texttt{TwoProd}
algorithm by Dekker \cite{Dekker1971}, respectively.

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{DeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\widehat{r} = 1 \ominus s\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\widehat{b}_j^{(k)} = \left(
              r \otimes \widehat{b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{b}_{j + 1}^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

In general, for a sequence \(v_0, \ldots, v_n\) we'll use the notation
\(v = \left[\begin{array}{c c c} v_0 & \cdots &
    v_n\end{array}\right]^T\) to indicate the vector containing all of
the values.

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \State \(\cdb{1}_j^{(n)} = 0\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \State \(\widehat{\ell}_{1, j}^{(k)} = \pi_1 \oplus \pi_2 \oplus
              \sigma_3 \oplus \left(\rho \otimes
              \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\cdb{1}_j^{(k)} =
              \widehat{\ell}_{1, j}^{(k)} \oplus
              \left(s \otimes \cdb{1}_{j + 1}^{(k + 1)}
              \right) \oplus
              \left(\widehat{r} \otimes
              \cdb{1}_j^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \widehat{b}_0^{(0)} \oplus
          \cdb{1}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{CompDeCasteljauK}\)}{$b, s, K$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \For{\(F = 1, \ldots, K - 1\)}
          \State \(\cdb{F}_j^{(n)} = 0\)
        \EndFor
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \\
          \State \(e = \left[\pi_1, \pi_2, \sigma_3\right]\)
          \State \(\delta b = \widehat{b}_j^{(k + 1)}\)
          \\
          \For{\(F = 1, \ldots, K - 2\)}
            \State \(\left[\eta, \widehat{\ell}\right] =
                \mathtt{LocalErrorEFT}(e, \rho, \delta b)\)
            \State \(L = \texttt{length}(\eta)\)
            \\
            \State \(\left[P_1, \eta_{L + 1}\right] = \mathtt{TwoProd}\left(
                s, \cdb{F}_{j + 1}^{(k + 1)}\right)\)
            \State \(\left[S_2, \eta_{L + 2}\right] =
                \mathtt{TwoSum}\left(\widehat{\ell}, P_1\right)\)
            \State \(\left[P_3, \eta_{L + 3}\right] = \mathtt{TwoProd}\left(
                \widehat{r}, \cdb{F}_j^{(k + 1)}\right)\)
            \State \(\left[\cdb{F}_j^{(k)}, \eta_{L + 4}\right]
                = \mathtt{TwoSum}\left(S_2, P_3\right)\)
            \\
            \State \(e = \eta\)
            \State \(\delta b = \cdb{F}_j^{(k + 1)}\)
          \EndFor
          \\
          \State \(\widehat{\ell} =
                \mathtt{LocalError}(e, \rho, \delta b)\)
          \State \(\cdb{K - 1}_j^{(k)} =
              \widehat{\ell} \oplus
              \left(s \otimes \cdb{K - 1}_{j + 1}^{(k + 1)}
              \right) \oplus
              \left(\widehat{r} \otimes
              \cdb{K - 1}_j^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{res} = \mathtt{SumK}\left(\left[
        \widehat{b}_0^{(0)}, \ldots, \cdb{K - 1}_0^{(0)}\right], K\right)\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

\noindent Noting that \(\ell_{F, j}\) contains \(5F - 1\) terms, one can
show that \texttt{CompDeCasteljauK} (Algorithm~\ref{alg:k-comp-de-casteljau})
requires
\begin{equation}
(15K^2 + 11K - 34)T_n + 6K^2 - 11K + 11 =
\bigO{n^2 K^2}
\end{equation}
flops to evaluate a degree \(n\) polynomial, where \(T_n\) is the
\(n\)th triangular number. As a comparison, the non-compensated form of
de Casteljau requires \(3 T_n + 1\) flops. In total this will require
\((3K - 4)T_n\) uses of \texttt{TwoProd}. On hardware that supports
FMA, \texttt{TwoProdFMA} (Algorithm~\ref{alg:two-prod-fma}) can be used
instead, lowering the flop count by \(15(3K - 4)T_n\). Another way
to lower the total flop count is to just use
\(\widehat{b}_0^{(0)} \oplus \cdots \oplus \cdb{K - 1}_0^{(0)}\)
instead of \texttt{SumK}. This will reduce the total by
\(6(K - 1)^2\) flops, however the results produced are (empirically)
identical. This makes sense: the whole point of \texttt{SumK}
is to filter errors in a summation so that the final operation produces
a sum of the form \(v_1 \oplus \cdots \oplus v_K\) where each
term is smaller than the previous by a factor of \(\mach\). This
property is already satisfied for the \(\cdb{F}_0^{(0)}\) so in
practice the \(K\)-compensated summation is likely not needed.

@@

\begin{lemma}\label{lemma:ell-tilde}
The local errors \(\widetilde{\ell}_{F, j}\) satisfy:
\begin{align}
\widetilde{\ell}_{1, j}^{(k)} &\leq
  \gamma_3 \left(
  (1 - s) \left|\widehat{b}_j^{(k + 1)}\right| +
  s \left|\widehat{b}_{j + 1}^{(k + 1)}\right|\right)
  \label{ell-tilde-1} \\
\widetilde{\ell}_{F + 1, j}^{(k)} &\leq
  \gamma_3 \left(
  (1 - s) \left|\cdb{F}_j^{(k + 1)}\right| +
  s \left|\cdb{F}_{j + 1}^{(k + 1)}\right|\right) +
  \gamma_{5F} \cdot \widetilde{\ell}_{F, j}^{(k)}
  \text{ for } F \geq 1.
\end{align}
\end{lemma}

@@

\begin{lemma}\label{lemma:L-and-D-bounds}
For \(s \in \left[0, 1\right]\), the Bernstein-type sum defined above
satisfies the following bounds:
\begin{align}
L_{F, n - k} &\leq \left[\left(3^F \binom{k}{F - 1} + \bigO{k^{F - 1}}\right)
  \mach^F + \bigO{\mach^{F + 1}}\right] \cdot \widetilde{p}(s) \\
\sum_{k = 0}^{n - 1} \gamma_{3k + 5F} L_{F, k} &\leq
  \left[\left(3^{F + 1} \binom{n}{F + 1} + \bigO{n^F}\right)
  \mach^{F + 1} + \bigO{\mach^{F + 2}}\right] \cdot \widetilde{p}(s).
  \label{L-sum-bound}
\end{align}
In particular, this means that
\(\sum_{k = 0}^{n - 1} \gamma_{3k + 5F} L_{F, k} =
\bigO{(3 n \mach)^{F + 1}} \cdot \widetilde{p}(s)\).
\end{lemma}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{VecSum}\)}{$p$}
      \State \(n = \texttt{length}(p)\)
      \For{\(j = 2, \ldots, n\)}
        \State \(\left[p_j, p_{j - 1}\right] = \mathtt{TwoSum}\left(
            p_j, p_{j - 1}\right)\)
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{res} = \mathtt{SumK}\)}{$p, K$}
      \For{\(j = 1, \ldots, K - 1\)}
        \State \(p = \mathtt{VecSum}(p)\)
      \EndFor
      \State \(\mathtt{res} = p_1 \oplus p_2 \oplus \cdots \oplus p_n\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

>>>>

The paper is organized as follows. Section~\ref{sec:notation} establishes
notation for error analysis with floating point operations, reviews
results about error-free transformations and reviews the
de Casteljau algorithm. In Section~\ref{sec:compensated-2},
the compensated algorithm for polynomial evaluation from \cite{Jiang2010} is
reviewed and notation is established for the expansion. In
Section~\ref{sec:compensated-k}, the \(K\) compensated algorithm is provided
and a forward error analysis is performed. Finally, in
Section~\ref{sec:numerical} we perform two numerical experiments to
give practical examples of the theoretical error bounds.

@@

An error-free transformation is a computational method where both
the computed result and the round-off error are returned. It
is considered ``free'' of error if the round-off can be represented
exactly as an element or elements of \(\mathbf{F}\).
The error-free transformations used in this paper are
the \texttt{TwoSum} algorithm by Knuth \cite{Knuth1997} and \texttt{TwoProd}
algorithm by Dekker \cite{Dekker1971}, respectively.

@@

  \begin{algorithmic}
    \Function{\(\mathtt{result} = \mathtt{DeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\widehat{r} = 1 \ominus s\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\widehat{b}_j^{(k)} = \left(
              \widehat{r} \otimes \widehat{b}_j^{(k + 1)}\right) \oplus
              \left(s \otimes \widehat{b}_{j + 1}^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{result} = \widehat{b}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

In general, for a sequence \(v_0, \ldots, v_n\) we'll refer to \(v\)
as the vector containing all of the values:
\(v = \left[\begin{array}{c c c} v_0 & \cdots &
    v_n\end{array}\right]^T.\)

@@

  \begin{algorithmic}
    \Function{\(\mathtt{result} = \mathtt{CompDeCasteljau}\)}{$b, s$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \State \(\cdb{1}_j^{(n)} = 0\)
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \State \(\widehat{\ell}_{1, j}^{(k)} = \pi_1 \oplus \pi_2 \oplus
              \sigma_3 \oplus \left(\rho \otimes
              \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\cdb{1}_j^{(k)} =
              \widehat{\ell}_{1, j}^{(k)} \oplus
              \left(s \otimes \cdb{1}_{j + 1}^{(k + 1)}
              \right) \oplus
              \left(\widehat{r} \otimes
              \cdb{1}_j^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{result} = \widehat{b}_0^{(0)} \oplus
          \cdb{1}_0^{(0)}\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{result} = \mathtt{CompDeCasteljauK}\)}{$b, s, K$}
      \State \(n = \texttt{length}(b) - 1\)
      \State \(\left[\widehat{r}, \rho\right] = \mathtt{TwoSum}(1, -s)\)
      \\
      \For{\(j = 0, \ldots, n\)}
        \State \(\widehat{b}_j^{(n)} = b_j\)
        \For{\(F = 1, \ldots, K - 1\)}
          \State \(\cdb{F}_j^{(n)} = 0\)
        \EndFor
      \EndFor
      \\
      \For{\(k = n - 1, \ldots, 0\)}
        \For{\(j = 0, \ldots, k\)}
          \State \(\left[P_1, \pi_1\right] = \mathtt{TwoProd}\left(
              \widehat{r}, \widehat{b}_j^{(k + 1)}\right)\)
          \State \(\left[P_2, \pi_2\right] = \mathtt{TwoProd}\left(
              s, \widehat{b}_{j + 1}^{(k + 1)}\right)\)
          \State \(\left[\widehat{b}_j^{(k)}, \sigma_3\right] =
              \mathtt{TwoSum}(P_1, P_2)\)
          \\
          \State \(e = \left[\pi_1, \pi_2, \sigma_3\right]\)
          \State \(\delta b = \widehat{b}_j^{(k + 1)}\)
          \\
          \For{\(F = 1, \ldots, K - 2\)}
            \State \(\left[\eta, \widehat{\ell}\right] =
                \mathtt{LocalErrorEFT}(e, \rho, \delta b)\)
            \State \(L = \texttt{length}(\eta)\)
            \\
            \State \(\left[P_1, \eta_{L + 1}\right] = \mathtt{TwoProd}\left(
                s, \cdb{F}_{j + 1}^{(k + 1)}\right)\)
            \State \(\left[S_2, \eta_{L + 2}\right] =
                \mathtt{TwoSum}\left(\widehat{\ell}, P_1\right)\)
            \State \(\left[P_3, \eta_{L + 3}\right] = \mathtt{TwoProd}\left(
                \widehat{r}, \cdb{F}_j^{(k + 1)}\right)\)
            \State \(\left[\cdb{F}_j^{(k)}, \eta_{L + 4}\right]
                = \mathtt{TwoSum}\left(S_2, P_3\right)\)
            \\
            \State \(e = \eta\)
            \State \(\delta b = \cdb{F}_j^{(k + 1)}\)
          \EndFor
          \\
          \State \(\widehat{\ell} =
                \mathtt{LocalError}(e, \rho, \delta b)\)
          \State \(\cdb{K - 1}_j^{(k)} =
              \widehat{\ell} \oplus
              \left(s \otimes \cdb{K - 1}_{j + 1}^{(k + 1)}
              \right) \oplus
              \left(\widehat{r} \otimes
              \cdb{K - 1}_j^{(k + 1)}\right)\)
        \EndFor
      \EndFor
      \\
      \State \(\mathtt{result} = \mathtt{SumK}\left(\left[
        \widehat{b}_0^{(0)}, \ldots, \cdb{K - 1}_0^{(0)}\right], K\right)\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

\noindent Noting that \(\ell_{F, j}\) contains \(5F - 1\) terms, one can
show that \texttt{CompDeCasteljauK} (Algorithm~\ref{alg:k-comp-de-casteljau})
requires
\begin{equation}
(15K^2 + 11K - 34)T_n + 6K^2 - 11K + 11 =
\bigO{n^2 K^2}
\end{equation}
flops to evaluate a degree \(n\) polynomial, where \(T_n\) is the
\(n\)th triangular number. As a comparison, the non-compensated form of
de Casteljau requires \(3 T_n + 1\) flops. In total this will require
\((3K - 4)T_n\) uses of \texttt{TwoProd}. On hardware that supports
FMA, \texttt{TwoProdFMA} (Algorithm~\ref{alg:two-prod-fma}) can be used
instead, lowering the flop count by \(15(3K - 4)T_n\). Another way
to lower the total flop count is to just use
\(\widehat{b}_0^{(0)} \oplus \cdots \oplus \cdb{K - 1}_0^{(0)}\)
instead of \texttt{SumK}; this will reduce the total by
\(6(K - 1)^2\) flops. When using a standard sum, the results produced
are (empirically) identical to those with \texttt{SumK}. This makes
sense: the whole point of \texttt{SumK}
is to filter errors in a summation so that the final operation produces
a sum of the form \(v_1 \oplus \cdots \oplus v_K\) where each
term is smaller than the previous by a factor of \(\mach\). This
property is already satisfied for the \(\cdb{F}_0^{(0)}\) so in
practice the \(K\)-compensated summation is likely not needed.

@@

\begin{lemma}\label{lemma:ell-tilde}
The local error bounds \(\widetilde{\ell}_{F, j}\) satisfy:
\begin{align}
\widetilde{\ell}_{1, j}^{(k)} &\leq
  \gamma_3 \left(
  (1 - s) \left|\widehat{b}_j^{(k + 1)}\right| +
  s \left|\widehat{b}_{j + 1}^{(k + 1)}\right|\right)
  \label{ell-tilde-1} \\
\widetilde{\ell}_{F + 1, j}^{(k)} &\leq
  \gamma_3 \left(
  (1 - s) \left|\cdb{F}_j^{(k + 1)}\right| +
  s \left|\cdb{F}_{j + 1}^{(k + 1)}\right|\right) +
  \gamma_{5F} \cdot \widetilde{\ell}_{F, j}^{(k)}
  \text{ for } F \geq 1.
\end{align}
\end{lemma}

@@

\begin{lemma}\label{lemma:L-and-D-bounds}
For \(s \in \left[0, 1\right]\), the Bernstein-type error sum defined above
satisfies the following bounds:
\begin{align}
L_{F, n - k} &\leq \left[\left(3^F \binom{k}{F - 1} + \bigO{k^{F - 1}}\right)
  \mach^F + \bigO{\mach^{F + 1}}\right] \cdot \widetilde{p}(s) \\
\sum_{k = 0}^{n - 1} \gamma_{3k + 5F} L_{F, k} &\leq
  \left[\left(3^{F + 1} \binom{n}{F + 1} + \bigO{n^F}\right)
  \mach^{F + 1} + \bigO{\mach^{F + 2}}\right] \cdot \widetilde{p}(s).
  \label{L-sum-bound}
\end{align}
In particular, this means that
\(\sum_{k = 0}^{n - 1} \gamma_{3k + 5F} L_{F, k} =
\bigO{(3 n \mach)^{F + 1}} \cdot \widetilde{p}(s)\).
\end{lemma}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{VecSum}\)}{$p$}
      \State \(n = \texttt{length}(p)\)
      \For{\(j = 2, \ldots, n\)}
        \State \(\left[p_j, p_{j - 1}\right] = \mathtt{TwoSum}\left(
            p_j, p_{j - 1}\right)\)
      \EndFor
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}

@@

  \begin{algorithmic}
    \Function{\(\mathtt{result} = \mathtt{SumK}\)}{$p, K$}
      \For{\(j = 1, \ldots, K - 1\)}
        \State \(p = \mathtt{VecSum}(p)\)
      \EndFor
      \State \(\mathtt{result} = p_1 \oplus p_2 \oplus \cdots \oplus p_n\)
    \EndFunction
  \end{algorithmic}
\end{breakablealgorithm}
